---
title: "hierarchical-clustering"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(stylo2gg)
```

In addition to two-dimensional relationships with principal components, stylo can also show a dendrogram for cluster analysis, showing texts' relationships based on their distance to each other. The resulting dendrogram shows texts in branches splitting off from one another at points that correspond to distance. In the orientation shown below, when two texts or groups of texts seem to have greater stylistic similarity, their branches will join nearer to the right side of the chart---nearer to a distance of zero. As the texts show greater dissimilarity, their branches will connect closer to the left side of the chart.

```{r eval=FALSE, include=FALSE}
# Only run this code chunk interactively, to create the needed files
federalist_mfw_hc <- stylo(gui = FALSE,
      corpus.dir = system.file("extdata/federalist", package = "stylo2gg"),
      custom.graph.title = "Federalist Papers",
      analysis.type = "CA",
      analyzed.features = "w",
      ngram.size = 1,
      display.on.screen = TRUE,
      sampling = "no.sampling",
      culling.max = 75,
      culling.min = 75,
      mfw.min = 900,
      mfw.max = 900,
      write.rds.file = TRUE)

saveRDS(federalist_mfw_hc, "federalist_mfw_hc.rds")
```

```{r eval=FALSE, echo=TRUE}
#| label: stylo_hc
#| cache: false
federalist_mfw_hc <- 
  stylo(gui = FALSE,
      corpus.dir = system.file("extdata/federalist", package = "stylo2gg"),
      custom.graph.title = "Federalist Papers",
      analysis.type = "CA",
      analyzed.features = "w",
      ngram.size = 1,
      display.on.screen = TRUE,
      sampling = "no.sampling",
      culling.max = 75,
      culling.min = 75,
      mfw.min = 900,
      mfw.max = 900)
```

```{r fig-stylo_hc2, echo=FALSE, fig.cap="Dendrogram of hierarchical clusters, prepared by stylo."}
#| fig.height: 10.0
# Actually run this code chunk, but don't show the code
federalist_mfw_hc <- readRDS("federalist_mfw_hc.rds")
readRDS("articles_CA_120_MFWs_Culled_75__Classic Delta_.rds")
```


Using this `federalist_mfw_hc` object, stylo2gg will create a similar cluster analysis:

```{r, fig.cap="To prepare dendrograms for display, stylo2gg uses <a href='https://cran.r-project.org/web/packages/dendextend/index.html'>dendextend</a>."}
#| fig.height: 10.0

federalist_mfw_hc |> 
  stylo2gg()
```

The addition of symbols to the chart allows for interpretability when printing, though these can be turned off with `shapes=FALSE`, as they can when showing PCA. In fact, many of the same options that apply to a visualization of principal components will also apply to cluster analysis:

```{r}
#| fig.height: 10.0
library(ggplot2)

federalist_mfw_hc |> 
  stylo2gg(viz = "CA",
           labeling = 2,
           highlight = 4,
           black = 3) + 
  theme(legend.position = "bottom") + 
  scale_shape_manual(values = 15:18)
```

### Options for distance and linkage

If they are available, stylo2gg will adopt stylo's settings for measuring distances and finding groups. If they're not available, it will default to finding distance using the Burrows delta (`distance.measure="delta"`) and grouping texts using the Ward1 method (`linkage="ward.D"`). Many other options native to stylo and to R can be set for these.

### Minimal display

When setting `viz="CA"` or importing directly from a stylo analysis with the option set `analysis.type="CA"`, stylo2gg will try to approximate the aesthetics shown by stylo, displaying an axis of distance and an explanatory caption. These options may not be desired since relative distances are already meaningful without the axis. To show a minimal dendrogram, choose the option `viz="hc"` (for hierarchical clustering)

```{r, fig.cap="This minimal dendrogram shows clusters using the Ward2 method based on Euclidean distances. If labels get clipped at the edge, consider expanding the limits using the added function shown here."}
#| fig.height: 10.0
federalist_mfw_hc |>
  stylo2gg(viz = "hc",
           shapes = FALSE,
           linkage = "ward.D2",
           distance = "euclidean") +
    expand_limits(y = 0)
```

### Highlighting results with boxes

As shown above, setting the `highlight` option to a category's number will draw a box around the branches containing this category. Unlike with principal components visualization, only one category can be highlighted at a time on a dendrogram; nevertheless, there are a few more options available to tweak this highlighting.

By default, stylo2gg will draw a single box, even if texts belonging to a category are distributed unevenly.

```{r, fig.cap="Drawing a highlight box when texts are distributed widely isn't very useful. The `highlight.nudge` option here moves the box's edge to avoid clipping labels; whether it is needed might depend on local installations."}
#| fig.height: 10.0
federalist_mfw_hc |>  
  stylo2gg(viz = "hc",
           shapes = FALSE,
           highlight = 1,
           highlight.nudge = 0.4,
           linkage = "ward.D2",
           distance = "euclidean") +
    expand_limits(y = 0)
```

Setting `highlight.single=FALSE` will instead draw multiple boxes while consolidating any contiguous branches:

```{r}
#| fig.height: 10.0
federalist_mfw_hc |> 
  stylo2gg(viz = "hc",
           shapes = FALSE,
           highlight = 1,
           highlight.nudge = 0.4,
           highlight.single = FALSE,
           linkage = "ward.D2",
           distance = "euclidean") +
    expand_limits(y = 0)
```

As an alternative to categorical highlighting on a dendrogram, stylo2gg can draw boxes indicated by each text's index number from the bottom, with the lowest text being number 1, the second-lowest text number 2, etc:

```{r, fig.cap="Manually indicate groupings to highlight using the `highlight.box` option. To aid in the counting, set `count.labels=TRUE`."}
#| fig.height: 10.0
federalist_mfw_hc |> 
  stylo2gg(viz = "hc",
           shapes = FALSE,
           highlight.box = c(6:19, 23:24, 29:63, 84:85),
           highlight.nudge = 0.4,
           linkage = "ward.D2",
           distance = "euclidean",
           show.zero = FALSE)
```

### Vertical dendrograms

Finally, set `horiz=FALSE` to display the cluster analysis vertically instead of horizontally:

```{r}
#| fig.width: 11.0
federalist_mfw_hc |> 
  stylo2gg(viz = "CA", horiz = FALSE) +
  theme(legend.position = "top")
```
